Making Sequential Programs Deterministic

### 2019-03-28 design.tex
More macros

@@

    \caption{Type checking rules}
    \label{typecheck-rules}
\end{figure}

@@

The core of the determinism type system is the following type qualifiers (see Figure~\ref{fig-determinism-hierarchy}):
%\todo{introduce just @Det and @NonDet first, and then @OrderNonDet after polymorphism.}
\begin{itemize}
    \item \<NonDet> indicates
    that the expression might have different values in two different executions.
    \item \<OrderNonDet> indicates that the expression is a collection or
      map (dictionary or associative array) that
      will have the same elements in every execution, but in a possibly different order.
    \item \<Det> indicates that
    the expression evaluates to the same value (with respect to \<.equals()>) in all
    executions; for a collection, iteration also yields the values in the same
    order.
\end{itemize}

@@

Our type system largely treats a map as a collection of key--value pairs.  Like any
collection, a map may be \<Det>, \<OrderNonDet>, or \<NonDet>.

@@

    $\infer[\textsc{ordernondet collection}]{\wellformed{\kappa_c \  \beta_c \langle\kappa_e \ \beta_e\rangle}}{\kappa_c : \|OrderNonDet/Det| & \kappa_e \sqsubseteq \kappa_c & \beta_c \sqsubseteq \<Collection>}$

@@

    $\infer[\textsc{nondet collection}]{\wellformed{\|NonDet| \ \beta_c\langle\|NonDet| \ \beta_e\rangle}}{\beta_c \sqsubseteq \<Collection>}$

@@

    $\infer[\textsc{ordernondet map}]{\wellformed{\kappa_m \  \beta_m \langle\kappa_k \ \beta_k, \kappa_v \ \beta_v\rangle}}{\kappa_m: \|OrderNonDet/Det| & \kappa_k \sqsubseteq \kappa_m & \kappa_v \sqsubseteq \kappa_m & \beta_m \sqsubseteq \<Map>}$

@@

    $\infer[\textsc{nondet map}]{\wellformed{\|NonDet| \ \beta_m\langle\|NonDet| \ \beta_k, \|NonDet| \ \beta_v\rangle}}{\beta_m \sqsubseteq \<Map>}$

@@

\todo{The collection rules do not permit nested collections; similarly, the
  (currently commented out) array rules did not permit any two-dimensional
  array to be well-formed.  There is a bigger issue here.
  Consider the ``nondet collection'' rule.  $\beta_c$ is a class (we need
  to use a different symbol for it, probably $c$), but $\beta_e$ should be
  a full type, which might contain type qualifiers within it.  (Actually,
  $\|NonDet|\ \beta_e$ is a type, not $\beta_e$ itself.)  So, there should
  be an antecedent stating that $\|NonDet| \beta_e$ is well-formed, and
  for the ``ordernondet collection'' rule stating that $\kappa_e \beta_e$
  is well-formed.  Likewise, antecedents are needed for the map rules.
  Otherwise, these rules permit construction of $\<Map>\langle\tau'\rangle$
  where $\tau'$ is itself not well-defined.}

@@

\TheDeterminismChecker has additional type refinement rules for the sorting
methods such as \<List.sort>, \<Collections.sort>, and \<Arrays.sort>:
 as shown in Figure~\ref{fig-sorting}. 
\begin{figure}
%     $\infer[\textsc{list sort}]{\Gamma \vdash x : \|Det \ List|\langle\|Det| \ \beta\rangle}{\Gamma \vdash x : \|OrderNonDet \ List|\langle\|Det| \ \beta\rangle & \|x.sort()|}$
%     
%     \bigskip

    $\infer[\textsc{sort}]{\Gamma \vdash x : \|Det \ List|\langle\|Det| \ \beta\rangle}{\Gamma \vdash x : \|OrderNonDet \ List|\langle\|Det| \ \beta\rangle & \|Collections.sort(x)|}$

@@

\subsection{Determinism checker specific rules for precision}\label{precision}
We annotated the \codeid{equals()} method of the \<Set> class in the JDK as
\begin{verbatim}
PolyDet$\uparrow$ boolean equals(PolyDet Set<E> this, PolyDet Object o)
\end{verbatim}
While this is sound, it is imprecise in the case of \codeid{OrderNonDet Sets}. We added the following rule shown in Figure~\ref{fig-precision-rules} to our checker to improve precision.
\begin{figure}
    \todo{add line break}
    $\infer[\textsc{set precision}]{\Gamma \vdash \|z : Det|}{\Gamma \vdash \|x| : \|OrderNonDet| \ \|Set|\langle \kappa_{x} \ \beta_{x}\rangle, \|y| : \|OrderNonDet| \ \|Set|\langle \kappa_{y} \ \beta_{y}\rangle & \|z = x.equals(y)| & \langle \kappa_{x} \ \beta_{x}\rangle \nsqsubseteq \langle \|OrderNonDet \ List| \rangle & \langle \kappa_{y} \ \beta_{y}\rangle \nsqsubseteq \langle \|OrderNonDet \ List| \rangle }$

@@

    $\infer[\textsc{map precision}]{\Gamma \vdash \|z : Det|}{\Gamma \vdash \|x : OrderNonDet \ Map\langle Det \ \beta_k, \kappa_v\ \beta_v \rangle, y : Det \ \beta| & \|z = x.get(y)|}$

>>>>

    $\infer[\textsc{invariant type param}]{\|method|(\overline{a_i}) : \tau_r}{\tau_r \  \|method|(\overline{\tau_i \  p_i}) & \overline{a_i : \tau_i}}$

    \caption{A sampling of standard type checking rules.  Most standard rules are omitted
    for brevity.  $\sqsubseteq$ is overloaded for classes and types.}
    \label{typecheck-rules}
\end{figure}

@@

    \caption{Type checking rules.  $\sqsubseteq$ is overloaded for classes,
    basetypes, type qualifiers, and types.  Most standard rules are omitted
    for brevity.}
    \label{typecheck-rules}
\end{figure}

@@

The core of the determinism type system is the following type qualifiers (see Figure~\ref{fig-determinism-hierarchy}):
%\todo{introduce just @Det and @NonDet first, and then @OrderNonDet after polymorphism.}
\begin{itemize}
    \item \<NonDet> indicates
    that the expression might have different values in two different executions.
    \item \<OrderNonDet> indicates that the expression is a collection or
      map that contains the same elements in every execution, but possibly
      in a different order.
    \item \<Det> indicates that
    the expression evaluates to the same value (with respect to \<.equals()>) in all
    executions; for a collection, iteration also yields the values in the same
    order.
\end{itemize}

@@

A map is a dictionary or associative array, such as a hash table.
Our type system largely treats a map as a collection of key--value pairs.  Like any
collection, a map may be \<Det>, \<OrderNonDet>, or \<NonDet>.

@@

    $\infer[\textsc{ordernondet collection}]{\wellformed{\kappa_c \  \beta_c \angles{\kappa_e \ \beta_e}}}{\kappa_c : \|OrderNonDet/Det| & \kappa_e \sqsubseteq \kappa_c & \beta_c \sqsubseteq \<Collection>}$

@@

    $\infer[\textsc{nondet collection}]{\wellformed{\|NonDet| \ \beta_c\angles{\|NonDet| \ \beta_e}}}{\beta_c \sqsubseteq \<Collection>}$

@@

    $\infer[\textsc{ordernondet map}]{\wellformed{\kappa_m \  \beta_m \angles{\kappa_k \ \beta_k, \kappa_v \ \beta_v}}}{\kappa_m: \|OrderNonDet/Det| & \kappa_k \sqsubseteq \kappa_m & \kappa_v \sqsubseteq \kappa_m & \beta_m \sqsubseteq \<Map>}$

@@

    $\infer[\textsc{nondet map}]{\wellformed{\|NonDet| \ \beta_m\angles{\|NonDet| \ \beta_k, \|NonDet| \ \beta_v}}}{\beta_m \sqsubseteq \<Map>}$

@@

\todo{The collection rules do not permit nested collections; similarly, the
  (currently commented out) array rules did not permit any two-dimensional
  array to be well-formed.  There is a bigger issue here.
  Consider the ``nondet collection'' rule.  $\beta_c$ is a class (we need
  to use a different symbol for it, probably $c$), but $\beta_e$ should be
  a full type, which might contain type qualifiers within it.  (Actually,
  $\|NonDet|\ \beta_e$ is a type, not $\beta_e$ itself.)  So, there should
  be an antecedent stating that $\|NonDet| \beta_e$ is well-formed, and
  for the ``ordernondet collection'' rule stating that $\kappa_e \beta_e$
  is well-formed.  Likewise, antecedents are needed for the map rules.
  Otherwise, these rules permit construction of $\<Map>\angles{\tau'}$
  where $\tau'$ is itself not well-defined.}

@@

\TheDeterminismChecker has additional type refinement rules for the sorting
methods such as \<List.sort>, \<Collections.sort>, and \<Arrays.sort>:
 as shown in Figure~\ref{fig-sorting}. 
\begin{figure}
%     $\infer[\textsc{list sort}]{\Gamma \vdash x : \|Det \ List|\angles{\|Det| \ \beta}}{\Gamma \vdash x : \|OrderNonDet \ List|\angles{\|Det| \ \beta} & \|x.sort()|}$
%     
%     \bigskip

    $\infer[\textsc{sort}]{\Gamma \vdash x : \|Det \ List|\angles{\|Det| \ \beta}}{\Gamma \vdash x : \|OrderNonDet \ List|\angles{\|Det| \ \beta} & \|Collections.sort(x)|}$

@@

\subsection{Determinism checker specific rules for precision}\label{precision}
We annotated the \codeid{equals()} method of the \<Set> class in the JDK as
\begin{verbatim}
PolyDet$\uparrow$ boolean equals(PolyDet Set<E> this, PolyDet Object o)
\end{verbatim}
While this is sound, it is imprecise in the case of \codeid{OrderNonDet Sets}. We added the following rule shown in Figure~\ref{fig-precision-rules} to our checker to improve precision.
\begin{figure}
    \todo{add line break}
    $\infer[\textsc{set precision}]{\Gamma \vdash \|z : Det|}{\Gamma \vdash \|x| : \|OrderNonDet| \ \|Set|\angles{ \kappa_{x} \ \beta_{x}}, \|y| : \|OrderNonDet| \ \|Set|\angles{ \kappa_{y} \ \beta_{y}} & \|z = x.equals(y)| & \angles{ \kappa_{x} \ \beta_{x}} \nsqsubseteq \angles{ \|OrderNonDet \ List| } & \angles{ \kappa_{y} \ \beta_{y}} \nsqsubseteq \angles{ \|OrderNonDet \ List| } }$

@@

    $\infer[\textsc{map precision}]{\Gamma \vdash \|z : Det|}{\Gamma \vdash \|x : OrderNonDet \ Map\angles{ Det \ \beta_k, \kappa_v\ \beta_v }, y : Det \ \beta| & \|z = x.get(y)|}$
